Learning how to act: making good decisions with machine learning

### 2017-09-28 thesis/thesis_map.tex
adds section for bob, starting work on talk

@@

\chapter*{Acknowledgements}
\input{abstract}

%\chapter*{Acknowledgements}

@@

Causality is often linked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}

@@

\paragraph{Thesis overview} This thesis is divided into three key chapters: learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \S\ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \S\ref{sec:defining-causal-effects} and \S\ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \S\ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system by observing, rather than acting on it. This is the problem of observational causal inference.

@@

@@

Remarkably for models developed relatively independently in fields with very different approaches and problems, the models we have discussed can be nicely unified for interventional queries (those that can be expressed with the do-notation) \citep{Pearl2000}. If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables. In other words, the network is a causal Bayesian network. All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.  Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant. This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}). Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks. We can also connect this to counterfactuals via:

@@

It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}. The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks. For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate. Then use propensity scores \citep{Rosenbaum1983} to estimate the effect. If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations. The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation. These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}.

In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks. The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations. The Neyman-Rubin framework builds on the understanding of randomised experiments and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment. This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}. In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field. Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}. In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule3_original}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{augmented graph $G^{\hat{X}}$}
\label{fig:rule3_augmented}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
\node[main node, above left=of 2](5){$\hat{X}$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3)
    (5) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{$G_{\overline{X}}$}
\label{fig:rule3_after}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Example application of equation \ref{eq:Do3}. $(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$. We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
\label{fig:rule3}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$X$};
\node[ob, right=of 1](2){$Z$};
\node[ob, right=of 2](3){$Y$};
\node[lt,above=of 2](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid2_instrumental_variable}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[ob, above left=of 2](4){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1)
    (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
\label{fig:identify}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.55]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}

@@

Reinforcement learning describes the problem of an agent interacting with an enviroment, learning by observing the outcome of its actions, with the goal of maximising some reward. These problems, which also incorporate planning, are extremely difficult because the value of an action is generally not immediately clear. The state of the enviroment may evolve over time and according to previously selected actions. Actions available at future timesteps can depend on those taken in the past, and rewards may be obtained only after a long sequence of actions. This makes it extremely difficult for the agent to acurately attribute value to each chosen action along the path to obtaining a given reward.

We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems. Multi-armed bandit problems also descibe an agent aiming to maximise some reward by interacting with an enviroment. However, the reward is observed immediately after the action is selected and the enviroment is stateless. Future rewards generated by the  enviroment in response to the agents actions do not depend on the previous actions of the agent. These assumptions are a reasonable approximation in settings where we face the same decision repeately with respect to many independent units or individuals and we have a good proxy for the outcome. In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms. We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomized experiments and highlight connections between the observational and interventional approaches to causal inference.

@@

where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$. The expectation of this variant of regret is the same as for the standard version defined in equation \ref{eqn:raw_regret2}. It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions. Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret. In conclusion, when selecting measures of bandit performance, it is worth noting whether they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.

@@

The type of problem we are concerned with is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown. In each season the farmer has enough time and money to intervene and control at most one of these
variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry. Where there is no intervenion, the temperature, soil, and moisture vary naturally from season to season due to weather conditions. These are all observed along with the final crop yield at the end of each season. How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?

@@

The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action, and on which variables we receive post-action feedback. If we can (cheaply) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem reduces to a standard multi-armed bandit problem.

%This is formalised in theorem \ref{theorem:cb-intervene-all-parents}.

If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example. If we receive post-action feedback on additional nodes the problem can be more interesting. In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected. Consider again the model in figure \ref{fig:unify_frameworks}. The causal structure implies:

@@

>>>>

\usepackage{geometry}
 \geometry{
 a4paper,
 left=40mm,
 right=20mm,
 top=20mm,
 bottom=30mm,
 footskip=10mm
 }

\usepackage{setspace}

@@

\chapter*{Acknowledgements}

%There are many people without whose support this thesis would not have been possible. I would like to thank my supervisor Dr Cheng Song Ong,

% Dr Mark Reid, Dr Tiberio Ceatano and Prof. Robert Williamson.

%My brother Tor for many interesting discussions on bandits.

%To my daughters, Freya and Anouk,

% supervisors
% family (Tor)
% thesis boot camp
% scholarship

%I am grateful to have recieved funding from an Australian Postgraduate Award and NICTA/Data61 scholarship to support this research.

\onehalfspacing

\input{abstract}

@@

Causality is often linked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distill how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}

@@

\paragraph{Thesis overview} This thesis is divided into three key chapters: learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data, where the goal is to learn the outcome of an external intervention in a system from data obtained by observing it without control over which actions are selected. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \S\ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \S\ref{sec:defining-causal-effects} and \S\ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \S\ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system by observing, rather than acting on it. This is the problem of observational causal inference. The key feature that distinguishes observational from interventional data is that the learning agent does not control the action about which they are trying to learn.

@@

\paragraph{Granger causality}
A discussion of approaches to (observational) causal inference would not be complete without a mention of Granger causality, \citep{Granger1969}. The fundamental idea behind Granger causality is to leverage the assumption that the future does not cause the past to test the existence and direction of a causal link between two time series. The basic approach is to test, for a pair of time series variables $X$ and $Y$, if $Y_t \ci (X_{1},...,X_{t-1}) | (Y_{1},...,Y_{t-1})$ - that is if the history of $X$ helps to predict $Y$ given the history of $Y$. The original formulation considered only pairs of variables and linear causal relationships but recent work has generalised the key idea to multiple variables and non-linear relationships. Unlike the previous models we have discussed, Granger causality does not provide us with a means to specify our assumptions about the causal structure between variables. Rather it aims to infer the causal structure of a structural equation model from observational data - subject to some assumptions. I would categorise Granger causality as method for \emph{causal discovery} in time series data, see the discussion of causal discovery versus causal effect estimation in section \ref{sec:causal effect estimation}.

@@

Remarkably for models developed relatively independently in fields with very different approaches and problems, causal Bayesian networks, counterfactuals and structural equation models can be nicely unified for interventional queries (those that can be expressed with the do-notation) \citep{Pearl2000}. These queries, and the assumptions required to answer them, can be mapped between the frameworks in a straightforward way, allowing techniques developed within one framework to be immediately applied within another. If the network for a structural equation model is acyclic, that is if starting from any node and following edges in the direction of the arrows you cannot return to the starting point, then it implies a recursive factorisation of the joint distribution over its variables. In other words, the network is a causal Bayesian network. All of the results that apply to causal Bayesian networks also apply to acyclic structural equation models.  Taking an action that sets a variable to a specific value equates to replacing the equation for that variable with a constant. This corresponds to dropping a term in the factorisation and the truncated product formula (equation \ref{eq:truncatedproduct}). Thus, the interventional query $P(Y|do(X))$ is identical in these two frameworks. We can also connect this to counterfactuals via:

@@

It is possible to relax the independence of errors assumption for SEMs to correspond exactly with the form of equation (\ref{eq:weakignore}) without losing any of the power provided by d-separation and graphical identification rules \citep{Richardson2013}. The correspondence between the models for interventional queries (those that can be phrased using the do-notation) makes it straightforward to combine key results and algorithms developed within any of these frameworks. For example, you can draw a causal graphical network to determine if a problem is identifiable and which variables should be adjusted for to obtain an unbiased causal estimate. Then use propensity scores \citep{Rosenbaum1983} to estimate the effect. If non-parametric assumptions are insufficient for identification or lead to overly large uncertainties, you can specify additional assumptions by phrasing your model in terms of structural equations. The frameworks do differ when it comes to causal queries that involve joint or nested counterfactuals and cannot be expressed with the do-notation. These types of queries arise in the study of mediation \citep{Pearl2014,Imai2010a,VanderWeele2011} and in legal decisions, particularly on issues such as discrimination \citep{Pearl2000}. The graphical approach to representing causal knowledge can be extended to cover these types of questions via Single World Intervention Graphs \citep{Richardson2013}, which explicitly represent counterfactual variables in the graph.

In practice, differences in focus and approach between the fields in which each model dominates eclipse the actual differences in the frameworks. The work on causal graphical models \citep{Pearl2000,Sprites2000} focuses on asymptotic, non-parametric estimation and rigorous theoretical foundations. The Neyman-Rubin framework builds on the understanding of randomised experiments and generalises to quasi-experimental and observational settings, with a particular focus on non-random assignment to treatment. Treatment variables are typically discrete (often binary). This research emphasises estimation of average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \citep{Rosenbaum1983}. In economics, inferring causal effects from non-experimental data to support policy decisions is central to the field. Economists are often interested in more informative measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \citep{Heckman2008}. The central approach to estimation is regression - which naturally handles continuous variables while discrete variables are typically encoded as indicator variables. In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models. However, these differences are not fundamental to the frameworks. Functional assumptions can be specified on the conditional distributions of (causal) Bayesian networks, counterfactuals can readily represent continuous treatments (eg $\cf{Y}{x}$), and structural equation models can represent complex non-linear relationships between both continuous and discrete variables.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{original network, $G$}
\label{fig:rule3_original}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{augmented graph $G^{\hat{X}}$}
\label{fig:rule3_augmented}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
\node[main node, above left=of 2](5){$\hat{X}$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (4)
    (4) edge (3)
    (5) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{$G_{\overline{X}}$}
\label{fig:rule3_after}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,main node/.style={observed}]
 %nodes
\node[main node](1){$Z$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, below right=of 2](4){$W$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (3)
    (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Example application of equation \ref{eq:Do3}. $(Y \ci \hat{X}|W,Z) \implies \P{Y|do(X),W,Z} = \P{Y|W,Z}$. We have to condition on $Z$ because conditioning on $W$ blocks the path $\hat{X}-X-W-Y$ but opens $\hat{X}-X-Z-Y$.}
\label{fig:rule3}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$X$};
\node[ob, right=of 1](2){$Z$};
\node[ob, right=of 2](3){$Y$};
\node[lt,above=of 2](4){$U$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3)
    (4) edge (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\caption{}
\label{fig:id3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid2_instrumental_variable}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[lt](1){$U$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[ob, above left=of 2](4){$Z$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3)
    (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.29\textwidth}
\centering
\vspace*{.5cm}
\caption{}
\label{fig:nid3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.8cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1)
    (1) edge (3);
\end{tikzpicture} 
\end{subfigure}
\caption{Examples of identifiable and non-identifiable queries. In subfigures (a), (b) and (c) the causal query $\P{Y|do(X)}$ is identifiable. In subfigures (d), (e) and (f) it is not.}
\label{fig:identify}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.48]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}

@@

Reinforcement learning describes the problem of an agent interacting with an environment, learning by observing the outcome of its actions, with the goal of maximising some reward. These problems, which also incorporate planning, are extremely difficult because the value of an action is generally not immediately clear. The state of the environment may evolve over time and according to previously selected actions. Actions available at future time steps can depend on those taken in the past, and rewards may be obtained only after a long sequence of actions. This makes it extremely difficult for the agent to accurately attribute value to each chosen action along the path to obtaining a given reward.

We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems. Multi-armed bandit problems also describe an agent aiming to maximise some reward by interacting with an environment. However, the reward is observed immediately after the action is selected and the environment is stateless. Future rewards generated by the  environment in response to the agents actions do not depend on the previous actions of the agent. These assumptions are a reasonable approximation in settings where we face the same decision repeatedly with respect to many independent units or individuals and we have a good proxy for the outcome. In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms. We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomised experiments and highlight connections between the observational and interventional approaches to causal inference.

@@

where $N_i(T)$ is the number of times arm $i$ was played up to timestep $T$ and $\Delta_i$ is the degree to which arm $i$ is sub-optimal, $\Delta_i = \mu^* - \mu_i$. The expectation of this variant of regret is the same as for the  version defined in equation \ref{eqn:raw_regret2}. It depends only on the randomness of the reward distribution only indirectly through the number of times each action is selected, which in turn depends only on the marginal distributions. Furthermore, this quantity has already been analysed in existing work on the concentration of bandit regret, \citep{Audibert2007,Audibert2009exploration} as a more tractable proxy to the standard regret. In conclusion, when selecting measures of bandit performance, it is worth noting whether they rely on counterfactual assumptions and considering if these assumptions are justifiable or needed for the specific problem on interest.

@@

The type of problem we are concerned with is best illustrated with an example. Consider a farmer wishing to optimise the yield of her crop. She knows that crop yield is only affected by temperature, a particular soil nutrient, and moisture level but the precise effect of their combination is unknown. In each season the farmer has enough time and money to intervene and control at most one of these
variables: deploying shade or heat lamps will set the temperature to be low or high; the nutrient can be added or removed through a choice of fertiliser; and irrigation or rain-proof covers will keep the soil wet or dry. Where there is no intervention, the temperature, soil, and moisture vary naturally from season to season due to weather conditions. These are all observed along with the final crop yield at the end of each season. How might the farmer best experiment to identify the single, highest yielding intervention in a limited number of seasons?

@@

The causal bandit problem takes on characteristics of different bandit settings depending on the action-space $\actionspace$ and corresponding costs, which variables are observable prior to selecting an action, and on which variables we receive post-action feedback. If we can (at no cost) intervene on all of the parents of $Y$ simultaneously then any context or alternative actions are irrelevant, and the problem can be treated as a standard multi-armed bandit problem over the set of actions that fully specify the values for the parents of $Y$. If feedback is received only on the reward node $\vec{X}^o = \set{Y}$, as in the standard bandit setting, then the do-calculus can be applied to eliminate some actions immediately, before any experiments are performed and then a standard bandit algorithm can be run on the remaining actions, see figure \ref{fig:cholesterol_graph} as an example. If we receive post-action feedback on additional nodes the problem can be more interesting. In addition to being able to eliminate some actions prior to sampling any data as in the previous case, taking one action may give us some information on actions that were not selected. Consider again the model in figure \ref{fig:unify_frameworks}. The causal structure implies:

@@

\subsubsection{Combining interventional and observational data}

In the previous sections we have demonstrated that knowledge of the causal structure between variables allows the rewards of multiple actions to be estimated simultaneously in an online, interventional setting, leading to more rapid identification of the optimal arm. In many cases we may also have access to observational data; in other words data sampled from $\P{\vec{X},Y|do()}$. Under the assumptions we made for algorithm \ref{alg:general}, we can incorporate such data in a straightforward way. Let $N$ be the number of observational data points previously collected and $T$ again be the number of rounds in which we can intervene. Combining these two periods, the values for the parents of $Y$ are sampled from;

\eqn {
Q'(\parents{Y}(X)) = \frac{N}{N+T}\Pn{do()}{\parents{Y}(X)} + \frac{T}{N+T}\sum_{b \in \calA}{\eta_b \Pn{b}{\parents{Y}(X)}}
}
